---
title: 'Envisionbox Summer School: Multimodal tools'
author: "A. Paxton (University of Connecticut)"
date: "11 July 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this tutorial, we'll be using a subset of the 
[Santa Barbara Corpus of Spoken American English](https://www.linguistics.ucsb.edu/research/santa-barbara-corpus-spoken-american-english) 
(CC-BY-ND 3.0, John W. Du Bois). I've chosen a series of dyadic conversations
for our tutorials; all of those should already be downloaded and available if
you've cloned the Summer School repository:

* `TilburgMultiscaleSummerschool2026/Datasets/SantaBarbaraCorpus`
  * `./metadata/`: Directory of participant metadata.
    * `./TRN/`: Directory of transcripts, with filenames to match `.wav` files.
    * `./WAV/`: Directory of audio files, with filenames to match `.trn` files.

The entire corpus is a great resource for a variety of different kinds of
conversations, including with larger groups than just dyads. I encourage you
to look around more if you're interested in context effects of interaction.

In addition, this code will require you to have the pretrained `word2vec` vectors
from the Google News corpus (
[Apache License 2.0](https://code.google.com/archive/p/word2vec/)
). A zipped version of the `.bin` file is available in this Summer School repository:

* `TilburgMultiscaleSummerschool2026/tuesday-multimodal/scripts`
  * `GoogleNews-vectors-negative300.bin.gz`: Zipped pretrained vectors. 
  Make sure that you unzip it prior to running this code. (Once successfully unzipped,
  it will be merely `.bin`; it will not unzip further.)
  
**Date last modified**: 13 July 2026
**Code written by**: A. Paxton (University of Connecticut)

***

# Preliminaries

```{r prelim, message=FALSE}

# clear the workspace
rm(list=ls())

# load in the libraries we'll need
library(tidyverse) # for a lot of everything
library(vader)
library(ConversationAlign)
library(word2vec)
library(tuneR)
library(knitr)
library(lsa)

# set seed for reproducibility
set.seed(0713)

```
```{r import_correlation_transform_variable, echo=FALSE}

# # This appears to have been part of a package that isn't CRAN-available, so
# # I'm including it from here: https://rdrr.io/github/martinctc/surveytools/src/R/cor_to_df.R

#' Convert correlation matrix into a tidy data frame
#' 
#' This function assumes that the input into cor() uses variable names, and available
#' variable labels are not used.
#' If either of the `label_table` or `id` arguments are unused, `cor_to_df`
#' simply returns a tibble without label matching.
#' Use the `label_table` and `id` arguments to match up variable names
#'  with the variable labels.
#' 
#' @param cor_m Correlation matrix in the form of an output from cor().
#' @param label_table A two-column table with one column used for matching and the other for returning labels.
#' @param id A character vector specifying the name of the matching / id column in the label_table.
#' @export
cor_to_df <- function(cor_m, label_table = NULL, id = NULL){
  if(is.null(label_table) | is.null(id)){
    message("Returning a tibble without changing labels")
    cor_m %>%
      tibble::as_tibble(rownames = "cor_matrix") %>%
      return()
  } else {
    if(ncol(label_table)!=2){
      stop("Matching table must have an id column and a return column.")
    } else {
      matched_lab <- names(label_table)[names(label_table)!=id]
      
      cor_m %>%
        tibble::as_tibble(rownames = "cor_matrix") %>%
        dplyr::left_join(label_table,by=c("cor_matrix"=id)) %>%
        dplyr::select(cor_matrix,matched_lab,everything())-> output_df
      
      labels <- output_df[[matched_lab]]
      
      names(output_df) <- c("cor_matrix",matched_lab,labels)
      return(output_df)
    }
  }
}
```

```{r directories}

# specify directories that we'll need
path_chat_transcript = "../../Datasets/SantaBarbaraCorpus/CHAT"
path_wav_audio = "../../Datasets/SantaBarbaraCorpus/WAV"

# figure out the files we have
list_transcript_files = list.files(path_chat_transcript, 
                                   pattern="*.cha",
                                   ignore.case=TRUE,
                                   full.names=TRUE)
list_audio_files = list.files(path_wav_audio, 
                              pattern="*.wav",
                              ignore.case=TRUE,
                              full.names=TRUE)

```

```{r read_in_semantic_vectors}

# read in the word2vec vectors trained on the Google News corpus
word2vec_pretrained_model = read.word2vec("../scripts/GoogleNews-vectors-negative300.bin")

```

***

# Data loading

Next, we'll cycle through all of the CHAT-formatted transcripts and convert them
into a neat, usable format. We will be relying heavily on a tutorial from the
[Language Technology and Data Analysis Laboratory](https://ladal.edu.au/) about
[corpus linguistics in R](https://ladal.edu.au/tutorials/corpus_linguistics/corpus_linguistics.html).

```{r load-cha-files}

# loop over paths, load, collapse, and clean each file  
corpus = sapply(list_transcript_files, function(next_file) {  
  
  # read the files as one big variable
  read_file = scan(next_file, 
                   what = "char", 
                   sep = "", 
                   quote = "", 
                   quiet = TRUE, 
                   skipNul = TRUE)  
  
  # collapse and try to remove unnecessary whitespaces
  read_file = paste0(read_file, sep = " ", collapse = " ")  
  read_file = stringr::str_squish(read_file)  
})  

# continue corpus cleanup
corpus = sapply(corpus, function(raw_corpus) {  
  
  # clean up our single line of text per transcript
  processing_corpus = stringr::str_trim(raw_corpus, side = "both")
  processing_corpus = stringr::str_squish(processing_corpus)
  processing_corpus = paste0(processing_corpus, collapse = " ")
  
  # insert split marker before each tier/speaker token, then split  
  processing_corpus = strsplit(  
    gsub("([%|*][a-z|A-Z]{2,4}[0-9]{0,1}:)", "~~~\\1", processing_corpus),  
    "~~~"  
  )  
})  

```

We're still processing things, but let's take a look at what the corpus looks
like, this far in.

```{r processing_peek}

# what does it look like right now?
str(corpus[1:3])  

```
This might look like a mess, but it's a lot of rich data here. Corpus linguistics
takes transcription and annotation very seriously, with specific codes for
various kinds of behaviors (like breathing, turn timing, and precise timing of 
overlapping speech). This specific dataset uses the coding system proposed by
Du Bois and colleagues (1993), and more details about the system can be found in
the PDF in the repository for this analysis.

For the sake of this tutorial, we'll be ignoring most of the codes, but they're
there in case you want to explore them.

***

# Data preparation

## Create dataframe structure

We'll start by pulling out the conversation metadata for each conversation,
which is the first line of each conversation file in this corpus.

```{r header-info}

# extract header elements for each file
conv_metadata = sapply(corpus, function(conv) conv[1])  

```

Then, let's split apart our corpus string by identifying new speaker turns.

```{r content}

# split content from the raw corpus string
content = sapply(corpus, function(raw_corpus) {  
  
  # strip out the header info and re-collapse
  processed_corpus = raw_corpus[2:length(raw_corpus)]  
  processed_corpus = paste0(processed_corpus, collapse = " ")  
  
  # split along speaker turns
  processed_corpus = stringr::str_split(  
    stringr::str_replace_all(processed_corpus, "(\\*[A-Z])", "~~~\\1"),  
    "~~~"  
  )  
})  

# and then see what we've got
content[[1]][1:6]  

```

Now, let's weave the header info with the content info to make our dataframe.

```{r}

# prepare dataframe elements with many rows we should have for each conversation
elements   = sapply(content, length)  
files_rep  = rep(names(elements), elements)  
conv_metadata_r = rep(conv_metadata, elements)  
rawcontent = as.vector(unlist(content))  

# create our dataframe
sbc_df = data.frame(  
  id         = seq_along(rawcontent),  
  file      = files_rep,  
  metadata   = conv_metadata_r,  
  rawcontent = rawcontent  
)  

# take a peek
head(sbc_df)

```

It's looking great! Let's move on.

## Extract metadata

We'll start by making the metadata content a bit more intelligible.

```{r clean_metadata}

# separate metadata by delimiters
sbc_df = sbc_df %>%
  separate_wider_delim(cols = metadata,
                       delim = " @",
                       names = c("encoding",
                                 "start",
                                 "language",
                                 "participants",
                                 "options",
                                 "ID1",
                                 "ID2",
                                 "ID3_or_media",
                                 "ID4_or_not",
                                 "media_or_not",
                                 "comment1_or_not",
                                 "comment2_or_not"),
                       too_many = "merge",
                       too_few = "align_start") %>%
  
  # drop the variables we don't actually need
  dplyr::select(-ID1, -ID2, -encoding, -start, -language, -options,
                -ID3_or_media) %>%
  
  # let's try solving for problematic labels
  mutate(conv_label = if_else(grepl("Comment",ID4_or_not),
                              ID4_or_not,
                              if_else(grepl("Comment", media_or_not),
                                      media_or_not,
                                      comment1_or_not))) %>%
  mutate(context = if_else(grepl("Comment", comment2_or_not),
                           comment2_or_not,
                           if_else(grepl("Comment", comment1_or_not),
                                   comment1_or_not,
                                   if_else(grepl("Comment", media_or_not),
                                           media_or_not,
                                           NA)))) %>%
  
  # strip out comment preface
  mutate(context = str_remove_all(context,"Comment: ")) %>%
  mutate(conv_label = str_remove_all(conv_label,"Comment: ")) %>%
  
  # and drop unneeded metadata
  dplyr::select(-contains("or_not")) %>%
  
  # why not reorder, too?
  relocate(id, file, conv_label, context, participants, rawcontent) %>%
  
  # clean file name: keep only the base name between the last / and .cha  
  mutate(file = gsub(".*/(.*?)\\.cha", "\\1", file))

# how are we looking?
head(sbc_df)

```

## Sort out the speaker data

Now, let's turn our attention to the content column. We'll start by extracting
the speaker label.

```{r extract_speaker}

# prepare speaker column
sbc_df = sbc_df %>% ungroup() %>%
  
  # keep everything before the first colon and strip remaining non-word characters  
  mutate(  
    speaker = stringr::str_remove_all(rawcontent, ":.*"),  
    speaker = stringr::str_remove_all(speaker, "\\W")  
  ) %>%
  
  # remove speakers from the content
  mutate(
    utterance = stringr::str_remove_all(rawcontent, "%mor:.*"),  
    utterance = stringr::str_remove_all(utterance, "%.*"),  
    utterance = stringr::str_remove_all(utterance, "\\*\\w{2,6}:"),  
    utterance = stringr::str_squish(utterance)
  )

```

## Extract non-linguistic verbal behaviors

Now that's finished, let's extract what we want from the content. Remember,
there are a lot of text-like comments that are, in fact, specific codes and not
participant utterances. We need to be very careful in extracting this to remove
codes and retain only speech. 

At the same time, though, we'll keep some codes about non-linguistic speech
characteristics. Again, these codes can be found in the Du Bois et al. (1993)
PDF.

```{r extract_nonlinguistic_stuff}

sbc_df = sbc_df %>% ungroup() %>%
  
  # extract various non-language markers and create new columns for each
  mutate(laughter = str_count(utterance, "@") + str_count(utterance, "\\&=laugh")) %>%
  mutate(short_pauses = str_count(utterance, "\\(\\.\\)")) %>%
  mutate(medium_pauses = str_count(utterance, "\\(\\.\\.\\)")) %>%
  mutate(long_pauses = str_count(utterance, "\\.\\.\\.")) %>%
  mutate(elongating = str_count(utterance, "\\:") + str_count(utterance, "\\&=length"))

```

## Ensure only text remains and extract timing

``` {r}

sbc_df = sbc_df %>% ungroup() %>%
  
  # extract start time
  mutate(start_time = str_extract(utterance, "[[:digit:]]+(?=\\_)")) %>%
  
  # clean out codes
  dplyr::mutate(  
    utterance = stringr::str_remove_all(utterance,"\\&\\{l=[[:alpha:]]+"),
    utterance = stringr::str_remove_all(utterance,"\\&\\{l="),
    utterance = stringr::str_remove_all(utterance,"\\&=[_[[:alpha:]]]+"),
    utterance = stringr::str_remove_all(utterance,"\\&\\}l=[[:alpha:]]+"),
    utterance = stringr::str_remove_all(utterance, "⌉[[:digit:]]+"),
    utterance = stringr::str_remove_all(utterance, "⌈[[:digit:]]+"),
    utterance = stringr::str_remove_all(utterance, "⌊[[:digit:]]+"),
    utterance = stringr::str_remove_all(utterance, "⌋[[:digit:]]+"),
    utterance = stringr::str_remove_all(utterance, "[⌉⌈⌊⌋]"),
    utterance = stringr::str_remove_all(utterance, "\\(\\.+\\)"),
    utterance = stringr::str_remove_all(utterance, "\\+\\/\\."),
    utterance = stringr::str_remove_all(utterance, "\\&\\{l=@"),
    utterance = stringr::str_remove_all(utterance, "\\&\\}l=@"),
    utterance = stringr::str_remove_all(utterance, "\\:"),
    utterance = stringr::str_remove_all(utterance, "ʔ"),
    utterance = stringr::str_remove_all(utterance, "\\+..."),
    utterance = stringr::str_remove_all(utterance, "\\[\\% laugh\\]"),
    utterance = stringr::str_remove_all(utterance, "@"),
    utterance = stringr::str_remove_all(utterance, "[[:digit:]]+\\_"),
    utterance = stringr::str_remove_all(utterance, "\\&\\{n=SNAP"),
    utterance = stringr::str_squish(utterance),
    utterance = stringr::str_replace_all(utterance, "\\s+", " "),
    utterance = stringr::str_replace_all(utterance, "[[:blank:]]+", " "),
    utterance = stringr::str_remove_all(utterance, "\\["),
    utterance = stringr::str_remove_all(utterance, "@End")
  ) %>%
  
  # extract the end time of the turn
  mutate(end_time = stringi::stri_extract_last_regex(utterance, "\\d+")) %>%
  
  # finish cleaning up the utterance, now that it's done
  mutate(
    utterance = stringr::str_remove_all(utterance, "[[:digit:]]+"),
    utterance = stringr::str_replace_all(utterance, "[^[:alpha:][:punct:]]", " "),
    utterance = stringr::str_remove_all(utterance, "\\.")) %>%
  
  # remove what we don't need
  dplyr::select(-rawcontent) %>%
  
  # filter out any turns with an NA start time
  dplyr::filter(!is.na(start_time))

# how's it look?
head(sbc_df)

```

## Concatentate adjacent turns

```{r compress_turns}

# figure out which turns have the same speaker
sbc_df = sbc_df %>%
  group_by(conv_label) %>%
  mutate(turn = row_number()) %>%
  mutate(last_speaker = lag(speaker)) %>%
  mutate(same_as_last = ifelse((last_speaker == speaker),
                               1,
                               0)) %>%
  mutate(revised_turn = ifelse(same_as_last == 1,
                               NA,
                               turn)) %>%
  mutate(revised_turn = ifelse(turn==min(turn, na.rm=TRUE),
                               1,
                               revised_turn)) %>%
  fill(., revised_turn, 
       .direction = "down") %>%
  ungroup() %>%
  
  # compress utterances and take unique values of everything else
  group_by(conv_label, speaker, revised_turn) %>%
  mutate(utterance = paste0(utterance, collapse = " ")) %>%
  mutate(start_time = min(start_time),
         end_time = max(end_time),
         laughter = sum(laughter),
         short_pauses = sum(short_pauses),
         medium_pauses = sum(medium_pauses),
         long_pauses = sum(long_pauses),
         elongating = sum(elongating),
         id = min(id)
  ) %>%
  dplyr::select(-turn, -last_speaker, -same_as_last) %>%
  unique() %>%
  
  # ditch revised turn and re-number
  arrange(conv_label, revised_turn) %>%
  mutate(turn = row_number()) %>%
  ungroup() %>%
  dplyr::select(-revised_turn) %>%
  relocate(id, file, conv_label, context, participants, speaker, utterance) 

# again, let's see how it looks
head(sbc_df)

```

I hope you've noticed by now how often we're checking our data. Especially for
complex, multiscale, multimodal data (but really for all data), it's imperative
to stay in touch with your data.

***

# Operationalizations and extractions

In this section, we're going to take our nicely cleaned data and start to
get at the core theoretical ideas we want to analyze. Thoughtful operationalization
is crucial, especially when you're working with large amounts of data
that can be sliced and quantified in a variety of ways. Multimodal and multiscale
data bring with them a lot of opportunities, but you're going to want to be
*painfully* clear about what you decided to do *a priori* (that is, **before** 
you started digging into the data) and *why* you chose to do that. As a domain
expert (or emerging domain expert), you could probably make a number of
arguments to explain why some quantification or another would be best suited
to your data *and* your question, and you want to be explicit about what you
intended to do from the outset.

This isn't to say that exploratory analyses are a bad thing. Exploratory analyses
can pave the way for new studies and can help pull apart nuances in the behavior
of interest. However, you *must* be clear (to yourself and your audiences) about
what was an an a priori hypothesis test and what was an exploratory analysis. 
The statistical and theoretical strength of a true a priori hypothesis is much
stronger than an exploratory analysis (see, e.g., 
[Wagenmakers et al., 2012](https://journals.sagepub.com/doi/10.1177/1745691612463078)).

With that in mind, we're planning some exploratory analyses here. For this 
tutorial, I tried to choose a range of settings from the Santa Barbara Corpus
that might show interesting dynamics:

```{r show_settings, include=FALSE, echo=TRUE}

unique(sbc_df$context)

```

## Prepare for alignment calculations with ConversationAlign

This appears to take a while, so we're going to let it cache results. This is
a very useful tool when you're actively developing the code, but *always*
remember to clear the cache *before* you start interpreting or presenting
your results.

```{r rename_vars_for_conversationalign, cache=TRUE}

# prepare for the ConversationAlign by renaming expected variables
sbc_df = sbc_df %>% ungroup() %>%
  dplyr::rename("Text_Raw" = utterance,
                "Participant_ID" = speaker,
                "Event_ID" = conv_label)

```

This next chunk needs to be run interactively, because it requires you to choose
which kinds of metrics you'd like to choose. I've chosen to include emotional
valence (`23`), log-scaled lexical frequency (`27`), and semantic diversity
(`40`). When you run this chunk interactively, you'll be able to see a whole lot
of other options that you can choose, but you can only choose up to 3 at a time.

```{r prepare_dyads, eval=FALSE}

# run preparation step to extract 
align_df = prep_dyads(sbc_df)

```

```{r save_align, include=FALSE, eval=TRUE}

# this is a secret chunk that will save the `align_df` dataframe for loading and rendering in RMarkdown
if (exists("align_df")){
  write.csv(x = align_df,
            file = "~/GitHub/TilburgMultiscaleSummerschool2026/tuesday-multimodal/data/align_df.csv",
            row.names = FALSE) 
}

```

```{r load_align, include=FALSE, eval=TRUE}

# this is a secret chunk that will load back in the `align_df` dataframe for loading and rendering
align_df = read.csv(file = "~/GitHub/TilburgMultiscaleSummerschool2026/tuesday-multimodal/data/align_df.csv",
                    header = TRUE)
                     
```

## Extract timing-related variables

Let's grab how long each turn was and how long it was from the end of the previous
turn to the start of the next turn. This means that negative turns were likely
interruptions; positive turns indicate a long gap in speech turns.

```{r extract_timing_variables}

align_df = align_df %>% ungroup() %>%
  
  # figure out how long each turn is
  mutate(end_time = as.numeric(end_time),
         start_time = as.numeric(start_time),
         turn_duration = end_time - start_time)

# extract how long elapsed between the end of one turn and the start of the next
inter_turn_intervals_df = align_df %>% ungroup() %>%
  dplyr::select(Event_ID, Turn_Count, start_time, end_time) %>%
  unique() %>%
  ungroup() %>%
  
  # and how long the breaks are between speakers' turns
  dplyr::group_by(Event_ID) %>%
  mutate(last_end = lag(end_time)) %>%
  dplyr::filter(!is.na(last_end)) %>%
  mutate(between_speaker_time = start_time - last_end) %>%
  ungroup()

# join with the main dataframe
align_df = dplyr::full_join(align_df, inter_turn_intervals_df,
                            by = join_by(Event_ID, Turn_Count, start_time, end_time))

```

In addition to inspecting the dataframe itself (which is good for language and 
harder for other modalities), you can also use plots to help understand what's going on.

```{r plot_timing_variables}

# what's our distribution of turn lengths?
ggplot(data = align_df,
       aes(x = turn_duration)) +
  geom_histogram() +
  ggtitle("Histogram of turn lengths")

# what's our distribution of onset-offset times?
ggplot(data = align_df,
       aes(x = between_speaker_time)) +
  geom_histogram() +
  ggtitle("Histogram of times between speaker turns")

```

Huh, we have some *very* far-off tails in these. See if you can figure out why!
Is it just a bug, or is there something happening in a corpus entry?

## Get turn-level aggregates across word norms

The `ConversationAlign` package calculates metrics at the (stemmed) lexical item
level, so if we want to see what a whole turns looks like, we need to do something
to unify them. You could make a case for some other method, but here, we're
going to go ahead and use mean (dropping `NA`).

```{r turn_level_norms}

turn_df = align_df %>% ungroup() %>%
  summarize(turn_length = n(),
            across(is.numeric|is.integer, ~ mean(., na.rm=TRUE)),
            .by = c(id, file, context, participants, Event_ID, Participant_ID, 
                    Exchange_Count, Turn_Count))

```

## Extract semantic vectors and similarities

We might also want to know the meaning (or semantic content) of what people are
saying in conversation. We could do that by annotating and hand-coding, but we
can also do that through vector space semantic analysis. There are lots of 
different ways of capturing this (starting back in the 1990s), but for now,
let's choose to use `word2vec`. 

For easier analysis, we're going to identify all unique words in our corpus,
look up the 300-dimensional vector for each in word2vec, and then re-constitute
using table joining. This will save us time and effort.

```{r extract_word2vec_dimensions}

# grab the unique words and their embeddings
unique_words = unique(align_df$Text_Prep)
embedding_matrix = matrix(NA,
                          nrow = length(unique_words),
                          ncol = 301)
for (i in c(1:length(unique_words))){
  next_word = unique_words[i]
  embedding_matrix[i,] = c(predict(word2vec_pretrained_model, 
                                   next_word, 
                                   type = "embedding"),
                           next_word)
}

# rename vector space variables in our matrix
embedding_matrix_df = data.frame(embedding_matrix) %>%
  rename(Text_Prep = X301) %>%
  rename_with(function(x) str_replace(x, "X","vector_"))

```

```{r reconstitute_word2vec_turns}

# re-embed the vectors into a whole space
vector_embeddings_df = dplyr::full_join(align_df, embedding_matrix_df,
                             by = c("Text_Prep")) %>%
  dplyr::mutate_at(vars(starts_with("vector_")), as.numeric) %>%
  dplyr::summarise(across(starts_with("vector_"),
                          function(x) mean(x, na.rm=TRUE)),
                   .by = c(Event_ID, Participant_ID, Exchange_Count, Turn_Count))

```

Now that we have our 300-dimensional vector, we can calculate the turn-to-turn
similarity by calculating the cosine similarity of the two turns. 

```{r extract_cosine_similarity, cache=TRUE}

# convert utterance-level to matrix
embedding_matrix = vector_embeddings_df %>% ungroup() %>%
  dplyr::select(starts_with("vector")) %>%
  mutate_all(as.numeric) %>%
  as.matrix(.)

# create utterance-to-utterance cosine similarity matrix
similarity_matrix = lsa::cosine(t(embedding_matrix))

# convert it to a longform dataframe 
similarity_df = cor_to_df(similarity_matrix) %>%
  pivot_longer(cols = starts_with("V"),
               names_to="turn_2",
               values_to="cos_similarity") %>%
  rename("turn_1" = cor_matrix) %>%
  mutate(turn_2 = str_remove_all(turn_2,"[[:alpha:]]"))

```

```{r label_similarity_matrix}

# extract conversation metadata for each turn
identifying_labels_df = vector_embeddings_df %>% ungroup() %>%
  rownames_to_column("rowname") %>%
  mutate(rowname = as.numeric(rowname)) %>%
  dplyr::select(-starts_with("vector"))

# merge with similarity
labeled_similarity_df = similarity_df %>% ungroup() %>%
  dplyr::mutate(turn_1 = as.numeric(turn_1),
                turn_2 = as.numeric(turn_2)) %>%
  dplyr::full_join(., identifying_labels_df,
                   by=c("turn_1"="rowname")) %>%
  dplyr::full_join(., 
                   identifying_labels_df,
                   by=c("turn_2"="rowname"),
                   suffix=c("_1","_2"))

```

Now that we've created a whole matrix of turn-to-turn similarity, we probably
want to do something useful with it. One useful thing (although not the only
useful thing) that we could do would be to see the average turn-to-turn similarity
in a specified window, excluding self-similarity. I'm arbitrarily choosing 3 windows,
but you could choose something else if you wanted.

```{r window_similarity}

# winnow to only include similarity for turns that are part of the same conversation
windowed_similarity_df = labeled_similarity_df %>% ungroup() %>%
  dplyr::filter(Event_ID_1 == Event_ID_2) %>%
  
  # remove same turns
  dplyr::filter(turn_1 != turn_2) %>%
  dplyr::filter(abs(turn_1-turn_2) <= 3) %>%
  
  # remove what we don't need
  select(-Event_ID_2, 
         -starts_with("Participant_"), 
         -starts_with("Exchange_"),
         -turn_2) %>%
  rename("Event_ID" = Event_ID_1,
         "Turn_Count" = Turn_Count_1) %>%
  
  # group to make the +/- 3-turn average 
  summarize(windowed_similarity = mean(cos_similarity, na.rm = TRUE),
            .by = c(Event_ID, Turn_Count))

```

```{r add_windowed_similarity}

# now, let's merge these values into the turn-level dataframe
turn_df = turn_df %>% ungroup() %>%
  dplyr::full_join(., windowed_similarity_df,
                   by = join_by(Event_ID, Turn_Count))

```

You'll probably guess what I'm about to suggest next. Let's take a look at our
data!

```{r plot_similarity_histogram, echo=FALSE}

ggplot(data = turn_df,
       aes(x = windowed_similarity)) +
  geom_histogram(bins = 30)

```

**Histogram of cosine similarity of turns (using 3-turn windows).**

## Identify fundamental frequency

I thought about adding a speech-based analysis here... but then I worried I was
already throwing too much into this tutorial. If you're interested, you could 
use the `start_turn` and `end_turn` numbers to identify the boundaries of each 
turn and then extract the F0 from that window. You could do any number of things 
with those data---like getting a slope for each person's F0 across the conversation, 
seeing whether the two co-vary, or even look at their turn-to-turn variability. This
[`wrassp` R package tutorial](https://ips-lmu.github.io/The-EMU-SDMS-Manual/chap-wrassp.html)
and 
[this Kaggle competition entry](https://www.kaggle.com/code/rtatman/time-aligned-spectrograms-with-chat-files-in-r)
would be good starting places.

***

# Data analysis

Congrats---we have our data ready to go! (Well, it's more than we actually need 
at this point, but we'll get to that later.) Now's a new fun part: trying out
permutation tests. 

Permutation tests might sound complicated, but don't worry. They are, truly,
one of the most straightforward hypothesis tests that you can do. Rather than
assuming that your data conform to some set of abstract distributions, your
data serve as their own baseline---even in the case of small or imbalanced
samples! 

The most difficult part of the permutation test is clearly specifying the null
hypothesis. It's a bit tricky, because null hypotheses something that many
behavioral scientists think of as being basic. In this case, the specifics of
your null hypothesis (and, conversely, your alternate hypothesis) are critical
to making sense of your data *and* getting useful results.

In this case, I'm going to focus on how **relationship context** (between
interlocutors) changes **language dynamics**, **turn-taking dynamics**, and **laughter**. 

* For the independent variable, I'm specifying "relationship context" as mutually
exclusive factors: 
friends (`SBC017`, `SBC044`), 
romantic partners (`SBC024`, `SBC034`, `SBC009`),
family (`SBC006`,`SBC047`, `SBC043`), and 
work (`SBC046`).
Notice that these are not balanced, nor are these the only divions that I *could*
have made. (For example, I was on the fence about an alternate division that 
would be more focused on topics of conversation, maybe with divisions for 
task-related conversations and personal experiences. But this is what I went with.)
* We'll be defining "laughter" as the turn-level presence or absence of laughter.
* We'll be defining "turn-taking dynamics" as the average turn overlap within a conversation.
* We'll be looking at "language dynamics" as the average windowed 3-turn similarity.

```{r label_categories}

analysis_df = turn_df %>% ungroup() %>%
  
  # label relationship context
  mutate(relationship = if_else(str_detect("SBC017|SBC044",
                                           file),
                                "friends",
                                if_else(str_detect("SBC024|SBC034|SBC009", 
                                                   file),
                                        "romantic",
                                        if_else(str_detect("SBC006|SBC047|SBC043", 
                                                           file),
                                                "family",
                                                if_else(str_detect("SBC046", 
                                                                   file),
                                                        "work",
                                                        NA))))) %>%
  
  # convert laughter to a binary
  mutate(laughter_binary = if_else(laughter == 0,
                                   0,
                                   1))

```

## Summarize dataset

Let's first see what our observed data look like.

```{r summarize_dataset}

# turns with laughter
describe_laughter_df = analysis_df %>% ungroup() %>%
  summarize(turns_with_laugh = sum(laughter_binary),
            .by = c(Event_ID, relationship))
head(describe_laughter_df)

# mean overlap of turns
describe_turn_overlap_df = analysis_df %>% ungroup() %>%
  summarize(average_overlap = mean(between_speaker_time, na.rm=TRUE),
            .by = c(Event_ID, relationship, Participant_ID))
head(describe_turn_overlap_df)

# mean windowed similarity
describe_windowed_coherence_df = analysis_df %>% ungroup() %>%
  summarize(average_semantic_coherence = mean(windowed_similarity, na.rm=TRUE),
            .by = c(Event_ID, relationship, Participant_ID))
head(describe_windowed_coherence_df)

```

If you look at the data, it turns out that we have some participants who
aren't in the target dyadic conversation! (Since these recordings are of
relatively "wild" conversations, there are other noises and people who were
captured.) Let's remove them before we continue.

```{r identify_nontarget_speakers}

# grab anyone who hasn't contributed more than 10 turns
observed_nontarget_speakers = analysis_df %>% ungroup() %>%
  summarize(total_turns = n(),
            .by = c(Event_ID, Participant_ID)) %>%
  dplyr::filter(total_turns <= 10) %>%
  dplyr::select(Participant_ID) %>%
  unique() %>%
  .$Participant_ID

# remove them from everything
analysis_df = analysis_df %>% ungroup() %>%
  dplyr::filter(!(Participant_ID %in% observed_nontarget_speakers))

```

But we need to summarize by relationship, not conversation.

```{r summarize_observed}

# turns with laughter
observed_laughter_df = analysis_df %>% ungroup() %>%
  summarize(observed_n = sum(laughter_binary),
            .by = c(relationship))

# turn overlap
observed_turn_overlap_df = analysis_df %>% ungroup() %>%
  summarize(observed_time = mean(between_speaker_time, na.rm=TRUE),
            .by = c(relationship))

# semantic coherence
observed_coherence_df = analysis_df %>% ungroup() %>%
  summarize(observed_coherence = mean(windowed_similarity, na.rm=TRUE),
            .by = c(relationship))

```

## Permutation tests

Each permutation test will require that we calculate it at the appropriate scale.
In other words, if our null hypothesis is at the level of the conversation versus
at the level of the participant, we will need to do different permutations. Rather
than re-run the permutation test each time, we're going to run them once, save them,
and re-use them each time we process.

Generally speaking, it's a best practice to randomize the independent variable,
rather the dependent variable. In our case, this means we're going to be
permuting along `relationship`.

```{r permutations}

# specify the total number of permutations
total_permutations = 2000

# specify where our baselines should live, if they exist
baseline_conversation_path = "../data/baseline_df.csv"

# load the permuted baselines, if we have them already
if (file.exists(baseline_conversation_path)){
  baseline_conversation_df = read.csv(baseline_conversation_path)
  
} else { # and if not, let's run them
  
  # subset only the variables we need from the analysis (to save on space)
  permutation_df = analysis_df %>% ungroup() %>%
    dplyr::select(Event_ID, Participant_ID, relationship, 
                  laughter_binary, between_speaker_time, windowed_similarity) 
  
  # cycle through our desired runs of permutations
  set.seed(0714)
  
  # let's make some containers
  baseline_conversation_df = data.frame()
  
  # and then cycle through everything
  for (permute_run in c(1:total_permutations)){
    
    # shuffle for relationship (between-dyads condition)
    next_conversation_permute = permutation_df %>% ungroup() %>%
      mutate(relationship = sample(relationship, length(relationship))) %>%
      mutate(run = permute_run) %>%
      mutate(permute_type = "conversation")
    baseline_conversation_df = rbind.data.frame(baseline_conversation_df,
                                                next_conversation_permute)
  }
  
  # add baseline designation
  baseline_conversation_df = baseline_conversation_df %>% ungroup() %>%
    mutate(data = "baseline")
  
  # write baseline dataframes to file
  write.csv(x = baseline_conversation_df,
            file = baseline_conversation_path,
            row.names = FALSE)
}
```

## Run significance tests

### Effects of relationship type on turns with laughter

```{r laughter_significance_tests}

# summarize number of turns with laughter by relationship for each run
laughter_baseline_df = baseline_conversation_df %>% ungroup() %>%
  dplyr::select(Event_ID, relationship, laughter_binary, data, run) %>%
  summarize(baseline_n = sum(laughter_binary),
            .by = c(relationship, run))

# see whether the data was higher or less than the baseline in each run
summary_laughter_comparison_df = dplyr::full_join(laughter_baseline_df, 
                                                  observed_laughter_df,
                                                  by = join_by(relationship)) %>% ungroup() %>%
  mutate_at(vars(observed_n, baseline_n), replace_na, 0) %>%
  mutate(diff = observed_n - baseline_n) %>%
  mutate(greater_than_chance = if_else(observed_n > baseline_n,
                                       1,
                                       0)) %>%
  mutate(less_than_chance = if_else(observed_n < baseline_n,
                                    1,
                                    0)) %>%
  mutate(at_chance = if_else(observed_n == baseline_n,
                             1,
                             0))

# extract summary statistics
test_laughter_stats_df = summary_laughter_comparison_df %>% ungroup() %>%
  dplyr::group_by(relationship) %>%
  dplyr::summarize(greater_than_chance = sum(greater_than_chance),
                   less_than_chance = sum(less_than_chance),
                   at_chance = sum(at_chance)) %>%

  rowwise() %>%
  mutate(total_obs = greater_than_chance + at_chance + less_than_chance) %>% 
  mutate(runs_without_combination = total_permutations - total_obs) %>%
  mutate(corrected_greater_than = greater_than_chance + runs_without_combination) %>%
  
  # to avoid p = 0 issues, we correct the "total observations" to include reality
  # (this isn't the best way to do it, but this is fairly conservative)
  dplyr::mutate(p_greater_than_chance = 1 - (corrected_greater_than/(total_permutations + 1)),
                p_less_than_chance = 1 - (less_than_chance/(total_permutations + 1))) 

# filter significant values with an alpha criterion of .05
significant_laughter_values_only = test_laughter_stats_df %>% ungroup() %>%
  dplyr::filter(p_greater_than_chance <= .05 |
                  p_less_than_chance <= .05)

# make some nicer tables
sig_laughter_key_df = test_laughter_stats_df %>% ungroup() %>%
  pivot_longer(cols = starts_with("p_"),
               names_to = "direction",
               values_to = "p_val") %>%
  dplyr::filter(p_val <= .05)
neat_sig_laughter_df = test_laughter_stats_df %>% ungroup() %>%
  mutate(color_cluster = relationship) %>%
  dplyr::filter(relationship %in% sig_laughter_key_df$relationship)

```

### Effects of relationship type on turn overlap

```{r overlap_significance_tests}

# summarize turn overlap by relationship for each run
overlap_baseline_df = baseline_conversation_df %>% ungroup() %>%
  dplyr::select(Event_ID, relationship, between_speaker_time, data, run) %>%
  summarize(baseline_time = mean(between_speaker_time, na.rm=TRUE),
            .by = c(relationship, run))

# see whether the data was higher or less than the baseline in each run
summary_overlap_comparison_df = dplyr::full_join(overlap_baseline_df, 
                                                  observed_turn_overlap_df,
                                                  by = join_by(relationship)) %>% ungroup() %>%
  mutate_at(vars(observed_time, baseline_time), replace_na, 0) %>%
  mutate(diff = observed_time - baseline_time) %>%
  mutate(greater_than_chance = if_else(observed_time > baseline_time,
                                       1,
                                       0)) %>%
  mutate(less_than_chance = if_else(observed_time < baseline_time,
                                    1,
                                    0)) %>%
  mutate(at_chance = if_else(observed_time == baseline_time,
                             1,
                             0))

# extract summary statistics
test_overlap_stats_df = summary_overlap_comparison_df %>% ungroup() %>%
  dplyr::group_by(relationship) %>%
  dplyr::summarize(greater_than_chance = sum(greater_than_chance),
                   less_than_chance = sum(less_than_chance),
                   at_chance = sum(at_chance)) %>%

  rowwise() %>%
  mutate(total_obs = greater_than_chance + at_chance + less_than_chance) %>% 
  mutate(runs_without_combination = total_permutations - total_obs) %>%
  mutate(corrected_greater_than = greater_than_chance + runs_without_combination) %>%
  
  # to avoid p = 0 issues, we correct the "total observations" to include reality
  # (this isn't the best way to do it, but this is fairly conservative)
  dplyr::mutate(p_greater_than_chance = 1 - (corrected_greater_than/(total_permutations + 1)),
                p_less_than_chance = 1 - (less_than_chance/(total_permutations + 1))) 

# filter significant values with an alpha criterion of .05
significant_overlap_values_only = test_overlap_stats_df %>% ungroup() %>%
  dplyr::filter(p_greater_than_chance <= .05 |
                  p_less_than_chance <= .05)

# make some nicer tables
sig_overlap_key_df = test_overlap_stats_df %>% ungroup() %>%
  pivot_longer(cols = starts_with("p_"),
               names_to = "direction",
               values_to = "p_val") %>%
  dplyr::filter(p_val <= .05)
neat_sig_overlap_df = test_overlap_stats_df %>% ungroup() %>%
  mutate(color_cluster = relationship) %>%
  dplyr::filter(relationship %in% sig_laughter_key_df$relationship)

```

### Effects of relationship type on semantic coherence

```{r coherence_significance_tests}

# summarize coherence overlap by relationship for each run
coherence_baseline_df = baseline_conversation_df %>% ungroup() %>%
  dplyr::select(Event_ID, relationship, windowed_similarity, data, run) %>%
  summarize(baseline_coherence = mean(windowed_similarity, na.rm=TRUE),
            .by = c(relationship, run))

# see whether the data was higher or less than the baseline in each run
summary_coherence_comparison_df = dplyr::full_join(coherence_baseline_df, 
                                                  observed_coherence_df,
                                                  by = join_by(relationship)) %>% ungroup() %>%
  mutate_at(vars(observed_coherence, baseline_coherence), replace_na, 0) %>%
  mutate(diff = observed_coherence - baseline_coherence) %>%
  mutate(greater_than_chance = if_else(observed_coherence > baseline_coherence,
                                       1,
                                       0)) %>%
  mutate(less_than_chance = if_else(observed_coherence < baseline_coherence,
                                    1,
                                    0)) %>%
  mutate(at_chance = if_else(observed_coherence == baseline_coherence,
                             1,
                             0))

# extract summary statistics
test_coherence_stats_df = summary_coherence_comparison_df %>% ungroup() %>%
  dplyr::group_by(relationship) %>%
  dplyr::summarize(greater_than_chance = sum(greater_than_chance),
                   less_than_chance = sum(less_than_chance),
                   at_chance = sum(at_chance)) %>%

  rowwise() %>%
  mutate(total_obs = greater_than_chance + at_chance + less_than_chance) %>% 
  mutate(runs_without_combination = total_permutations - total_obs) %>%
  mutate(corrected_greater_than = greater_than_chance + runs_without_combination) %>%
  
  # to avoid p = 0 issues (which are bad if we need to correct for multiple comparisons), 
  # we can correct the "total observations" to include reality in the denominator (+ 1)
  dplyr::mutate(p_greater_than_chance = 1 - (corrected_greater_than/(total_permutations + 1)),
                p_less_than_chance = 1 - (less_than_chance/(total_permutations + 1))) 

# filter significant values with an alpha criterion of .05
significant_coherence_values_only = test_coherence_stats_df %>% ungroup() %>%
  dplyr::filter(p_greater_than_chance <= .05 |
                  p_less_than_chance <= .05)

# make some nicer tables
sig_coherence_key_df = test_coherence_stats_df %>% ungroup() %>%
  pivot_longer(cols = starts_with("p_"),
               names_to = "direction",
               values_to = "p_val") %>%
  dplyr::filter(p_val <= .05)
neat_sig_coherence_df = test_coherence_stats_df %>% ungroup() %>%
  mutate(color_cluster = relationship) %>%
  dplyr::filter(relationship %in% sig_laughter_key_df$relationship)

```

## Results tables

```{r create_results_tables}

# laughter results table
readable_laughter_table = dplyr::left_join(neat_sig_laughter_df,
                                          observed_laughter_df,
                                          by = "relationship") %>%
  dplyr::select(relationship, observed_n, contains("p_"))

# turn overlap results table
readable_overlap_table = dplyr::left_join(neat_sig_overlap_df,
                                          observed_turn_overlap_df,
                                          by = "relationship") %>%
  dplyr::select(relationship, observed_time, contains("p_"))

# semantic coherence results table
readable_coherence_table = dplyr::left_join(neat_sig_coherence_df,
                                          observed_coherence_df,
                                          by = "relationship") %>%
  dplyr::select(relationship, observed_coherence, contains("p_"))

```

```{r show_results_tables, echo=FALSE}

kable(readable_overlap_table, 
      caption = paste0("**Results for turn overlap analysis.** Mean overlap ",
                  "in ms per conversation. Mean observed overlap by each ",
                  "relationship type included as `observed_time` variable. ",
                  "Only relationship types with significant values included."))

kable(readable_laughter_table, 
      caption = paste0("**Results for laughter analysis.** Number of turns ",
                  "with laughter per conversation. Total number of observed turns by each ",
                  "relationship type included as `observed_n` variable. ",
                  "Only relationship types with significant values included."))

kable(readable_laughter_table, 
      caption = paste0("**Results for windowed coherence analysis.** Mean semantic coherence ",
                  "in a +/- 3-turn window. Mean observed coherence by each ",
                  "relationship type included as `observed_coherence` variable. ",
                  "Only relationship types with significant values included."))
```

***

# Data visualization

We've already examined the significance values, but let's take a look at some
data visualizations. For these, we'll be showing the histogram of the entire
permuted dataset (faceted by relationship type) and a single vertical line (in red) 
to show where the observed data lie.

```{r laugh_histogram, echo = FALSE}

# histogram of permutation test baselines
ggplot(data = summary_laughter_comparison_df,
       aes(x = baseline_n,
           fill = relationship)) +
  facet_wrap(vars(relationship)) +
  stat_bin(bins = 30) +
  scale_fill_manual(breaks = c("family",
                               "friends",
                               "romantic",
                               "work"),
                    values = c("#F8766D",
                               "#D89000",
                               "#39B600",
                               "#00B0F6"),
                    guide = "none") +
  geom_vline(data = . %>% 
               group_by(relationship) %>% 
               summarise(value=mean(observed_n)), 
             aes(xintercept=value),
             color = "red") +
  labs(y = "Count",
       x = "Turns with laughter",
       title = "Turns with laughter by relationship type")

```

```{r overlap_histogram, echo = FALSE}

# histogram of permutation test baselines
ggplot(data = summary_overlap_comparison_df,
       aes(x = baseline_time,
           fill = relationship)) +
  facet_wrap(vars(relationship)) +
  stat_bin(bins = 30) +
  scale_fill_manual(breaks = c("family",
                               "friends",
                               "romantic",
                               "work"),
                    values = c("#F8766D",
                               "#D89000",
                               "#39B600",
                               "#00B0F6"),
                    guide = "none") +
  geom_vline(data = . %>% 
               group_by(relationship) %>% 
               summarise(value=mean(observed_time)), 
             aes(xintercept=value),
             color = "red") +
  labs(y = "Count",
       x = "Mean overlap between speakers' turns",
       title = "Turns overlap by relationship type")

```

```{r coherence_histogram, echo = FALSE}

# histogram of permutation test baselines
ggplot(data = summary_coherence_comparison_df,
       aes(x = baseline_coherence,
           fill = relationship)) +
  facet_wrap(vars(relationship)) +
  stat_bin(bins = 30) +
  scale_fill_manual(breaks = c("family",
                               "friends",
                               "romantic",
                               "work"),
                    values = c("#F8766D",
                               "#D89000",
                               "#39B600",
                               "#00B0F6"),
                    guide = "none") +
  geom_vline(data = . %>% 
               group_by(relationship) %>% 
               summarise(value=mean(observed_coherence)), 
             aes(xintercept=value),
             color = "red") +
  labs(y = "Count",
       x = "Mean windowed semantic similarity",
       title = "Windowed semantic coherence by relationship type")

```

These histograms---in addition to showing off the statistical significance---highlight
the sensitivity of permutation tests to the properties of the real behavior.
The number of turns in each type of conversation varied, for example, and that 
gets accounted for *if* we choose the appropriate level of permutation
for our null hypothesis. We don't have to wrestle with the appropriate linking
functions or nested intercepts; we get this respect for the data "for free"
with permutation tests.

Out of curiosity---does anything strike you as interesting or unexpected about
these results? If you read the transcripts (or, better yet, listen to the audio), 
do you feel as though these analyses capture something important or useful about
the conversations? Why or why not? What might you do differently?